---
title:        "README"
bibliography: inst/REFERENCES.bib
csl:          inst/apa.csl
output:       github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->
```{r include = FALSE, eval = FALSE}

```
```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.path = "man/figures/README-",
  dev      = "png"
)
if(.Platform$OS.type == "windows"){
  knitr::opts_chunk$set(dev.args = list(type = "cairo"))
}
```

<!-- badges: start -->
[![CRAN status](https://www.r-pkg.org/badges/version/RoBTT)](https://CRAN.R-project.org/package=RoBTT)
<!-- badges: end -->

<!-- 
#[![R-CRAN-check](https://github.com/FBartos/RoBTT/workflows/R-CMD-check/badge.svg)](https://github.com/FBartos/RoBTT/actions)
[![R-tests](https://github.com/FBartos/RoBTT/workflows/R-CMD-tests/badge.svg)](https://github.com/FBartos/RoBTT/actions)
#[![Codecov test coverage](https://codecov.io/gh/FBartos/RoBTT/branch/master/graph/badge.svg)](https://app.codecov.io/gh/FBartos/RoBTT?branch=master)
-->

# Robust Bayesian T-Test (RoBTT)

This package provides an implementation of Bayesian model-averaged t-tests that allows users to draw inference about the presence vs absence of the effect, heterogeneity of variances, and outliers. The RoBTT packages estimates model ensembles of models created as a combination of the competing hypotheses and uses Bayesian model-averaging to combine the models using posterior model probabilities. Users can obtain the model-averaged posterior distributions and inclusion Bayes factors which account for the uncertainty in the data generating process. User can define a wide range of informative priors for all parameters of interest. The package provides convenient functions for summary, visualizations, and fit diagnostics.

See our manuscripts for more information about the methodology:

  - @maier2022bayesian introduces a robust Bayesian t-test that model-averages over normal and t-distributions to account for the uncertainty about potential outliers,
  - @godmann2024how introduces a truncated Bayesian t-test that accounts for outlier exclusion when estimating the models.

We also prepared vignettes that illustrate functionality of the package:

 - [Introduction to RoBTT](https://fbartos.github.io/RoBTT/articles/Introduction_to_RoBTT.html)
 - [Truncated T-Tests](https://fbartos.github.io/RoBTT/articles/Truncated_t_test.html)


## Installation

The release version can be installed from CRAN:
``` r
install.packages("RoBTT")
```
and the development version of the package can be installed from GitHub:

``` r
devtools::install_github("FBartos/RoBTT")
```


### References
