---
title: "Truncated Bayesian Model-Averaged T-Test"
author: "Henrik R. Godmann & František Bartoš"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    self_contained: yes
bibliography: ../inst/REFERENCES.bib
csl: ../inst/apa.csl
vignette: >
  %\VignetteIndexEntry{Truncated Bayesian Model-Averaged T-Test}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown_notangle}
---


```{r setup, include=FALSE}
is_check <- ("CheckExEnv" %in% search()) ||
              any(c("_R_CHECK_TIMINGS_", "_R_CHECK_LICENSE_") %in% names(Sys.getenv())) ||
              !file.exists("../models/Truncated/fit1_trunc.RDS")
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  eval      = !is_check,
  dev      = "png")
if(.Platform$OS.type == "windows"){
  knitr::opts_chunk$set(dev.args = list(type = "cairo"))
}
```

```{r include = FALSE, eval = FALSE}
# pre-fit all models (easier to update the code on package update)
library(RoBTT)

set.seed(42)
x1 <- rnorm(100, 0, 1)
x2 <- rnorm(100, 0, 1)

stats1 <- boxplot.stats(x1)
lower_whisker1 <- stats1$stats[1]
upper_whisker1 <- stats1$stats[5]


stats2 <- boxplot.stats(x2)
lower_whisker2 <- stats2$stats[1]
upper_whisker2 <- stats2$stats[5]

# Exclude outliers based on identified whiskers
x1_filtered <- x1[x1 >= lower_whisker1 & x1 <= upper_whisker1]
x2_filtered <- x2[x2 >= lower_whisker2 & x2 <= upper_whisker2]

# Define whiskers for truncated likelihood application
whisker1 <- c(lower_whisker1, upper_whisker1)
whisker2 <- c(lower_whisker2, upper_whisker2)

fit1_trunc <- RoBTT(
  x1 = x1_filtered, x2 = x2_filtered,
  truncation = list(x1 = whisker1, x2 = whisker2),
  seed = 1, parallel = FALSE)

saveRDS(fit1_trunc, file = "../models/Truncated/fit1_trunc.RDS")
```

# Truncated Bayesian Model-Averaged T-Test

## Introduction
This vignettes accompanies our recent manuscript ''How to exclude outliers without biasing the Bayes factor'' [@godmann2024how] and shows how to use the `RoBTT` R package to estimate a truncated Bayesian model-averaged independent samples $t$-test (TrBTT). TrBTT adapts the t-test to researchers' outlier handling and thus mitigates the unwanted side effects of outlier exclusion on the inferences. For a general introduction to the RoBTT package, see the [Introduction to RoBTT](https://fbartos.github.io/RoBTT/articles/Introduction_to_RoBTT.html) vignette.

## Background
Outliers can lead to biased analysis results. However, the widely applied approach of simply excluding extreme observations without changing the analysis is also not appropriate, as it often leads to inflated evidence. This vignette introduces a truncated version of the Bayesian model-averaged independent samples $t$-test and demonstrates an alternative way of handling outliers in a Bayesian hypothesis testing framework. TrBTT incorporates the Bayesian model-averaging approach with a truncated likelihood. As such, TrBTT offers a robust solution for conducting independent samples $t$-tests that are less susceptible to the influence of outlier. 

The TrBTT truncates the likelihood identically to the truncation applied to data. As such, it overcomes the otherwise biased variance estimates due to outlier exclusion. It simultaneously model-averages across $4$ different models;

  1) model assuming no effect, and equal variances across group,
  2) model assuming no effect, and unequal variances across groups,
  3) model assuming presence of the effect, and equal variances across group,
  4) and model assuming presence of the effect, and unequal variances across groups.
  
For all models, the likelihood is adjusted according to the specified values. Inferences are based on a weighted average of each model's predictive performance.

## Application

### Installing and Loading RoBTT

First, we ensure that the RoBTT package is installed and loaded into the R session:

```{r, echo=T,error=F, message=F, warning=F, results="hide"}
# Install RoBTT from CRAN
# install.packages("RoBTT")

# Load the RoBTT package
library(RoBTT)
```

### Example Data Generation
We generate some example data to demonstrate the functionality of the test:

```{r, echo=T,error=F, message=F, warning=F, results="hide"}
set.seed(42)
x1 <- rnorm(100, 0, 1)
x2 <- rnorm(100, 0, 1)
```

### Model-Averaged Truncated Bayesian Independent Samples $t$-Test

#### 1. Manual Outlier Exclusion Based on Specific Cutoffs.
First, we demonstrate how to manually exclude outliers using specific cut-offs and then apply truncation to the likelihood function. It is possible to specify specific cut-offs for each group separately, for instance because the box plot method was used before to identify cut-offs. Further, it is possible to define a cut-off that was applied to both groups, for instance when all response times slower than $200$ ms and higher than $1000$ ms should be excluded in both groups.

First, a cut-off range for each group:

```{r, echo=T,error=F, message=F, warning=F, results="hide"}
# Identify outliers using boxplot statistics for each group
stats1 <- boxplot.stats(x1)
lower_whisker1 <- stats1$stats[1]
upper_whisker1 <- stats1$stats[5]


stats2 <- boxplot.stats(x2)
lower_whisker2 <- stats2$stats[1]
upper_whisker2 <- stats2$stats[5]

# Exclude outliers based on identified whiskers
x1_filtered <- x1[x1 >= lower_whisker1 & x1 <= upper_whisker1]
x2_filtered <- x2[x2 >= lower_whisker2 & x2 <= upper_whisker2]

# Define whiskers for truncated likelihood application
whisker1 <- c(lower_whisker1, upper_whisker1)
whisker2 <- c(lower_whisker2, upper_whisker2)
```

We can then fit the truncated RoBTT:

```r
# Fit the RoBTT model with truncation using the filtered data
fit1_trunc <- RoBTT(
  x1 = x1_filtered, x2 = x2_filtered,
  truncation = list(x1 = whisker1, x2 = whisker2),
  seed = 1, parallel = FALSE)
```

The following specification of the truncation argument implements the same cut-off range for both group:

```{r, echo=T,error=F, message=F, warning=F, results="hide"}
cut_off <- c(-2,2)

x1 <- x1[x1 >= -2 & x1 <= 2]
x2 <- x2[x2 >= -2 & x2 <= 2]

# fit RoBTT with truncated likelihood
fit1_trunc  <- RoBTT(
  x1 = x1, x2 = x2, 
  truncation = list(x = cut_off),
  seed = 1, parallel = FALSE)
```

We can retrieve a summary of the estimates with the code below. This function gives us the posterior mean and median estimate for the effect (Cohen’s d) and 95\% central credible interval. It further shows the estimates for mean and variance in each of the groups. In addition, it shows the inclusion Bayes factor for a presence of an effect across all models, as well as for including heterogeneity.

```{r}
summary(fit1_trunc, group_estimates = TRUE)
```

We can also look at the estimates for the individual models and retrieve a table including information about the prior distributions. The inclusion Bayes factor shows the evidence for the inclusion of the individual models.

```{r}
summary(fit1_trunc, group_estimates = TRUE, type= "models")
```


#### 2. Applying Direct Truncation Based on Standard Deviations
The RoBTT package also allows specifying truncation directly based on standard deviations, simplifying the process of outlier handling. The function will then proceed to exclude extreme observations and accordingly to truncate the likelihood, before estimating the model. No manual outlier exclusion is needed before. This is again possible for the same standard deviation value sigma to be applied to both groups, as well as to specify different standard deviations per group.

First, a cut-off range sigma for both groups:

```r
# Fit the model with direct truncation based on standard deviations
fit1_trunc <- RoBTT(
  x1 = x1, x2 = x2,
  truncation = list(sigma = 2.5),
  seed = 1, parallel = FALSE)
```

Second, a different standard deviation sigma for each group:

```r
# Fit the model with direct truncation based on standard deviations
fit1_trunc <- RoBTT(
  x1 = x1, x2 = x2,
  truncation = list(sigma1 = 2, sigma2 = 2.5),
  seed = 1, parallel = FALSE)
```

## Conclusions

This vignette demonstrated outlier handing with truncated Bayesian model-averaged t-test implemented in the `RoBTT` R package. For methodological background see @godmann2024how. 

### References
